---
title: "Exploration of wild-type and transgenic variants of Arabidopsis Thaliana"
author: "Yves Vernooy"
format: 
  dashboard:
    theme: slate
    orientation: columns
embed-resources: true
---

```{r imports}
#| message: false
#| output: false

# Note: Some of these libraries are really annoying to install and
# need different commands beyond the typical install.packages()
library(knitr)
library(dplyr)
library(tidyverse)
library(pheatmap)
library(GEOquery)
library(ggplot2)
library(DESeq2)

# read in data
getGEOSuppFiles("GSE277320")
counts <- read.delim("GSE277320/GSE277320_CountsMatrix.txt.gz", row.names = 1)
```

# Introduction

## Row {height=60%}

This project is a Quarto dashboard exploring differential gene expression 
in *Arabidopsis thaliana* under heat stress conditions, using a publicly 
available RNA-seq dataset from NCBI's Gene Expression Omnibus (GEO accession: 
GSE277320). This dataset was generated by Cabrales-Orona et al. (2026) and 
contains gene expression profiles from wild-type *Arabidopsis* plants and 
transgenic lines overexpressing RNA-binding proteins originally identified in 
grain amaranth (*Amaranthus hypochondriacus*), a heat-tolerant crop plant. 
All 30 samples were collected under heat shock conditions across five time 
points, with two biological replicates per group.

This project aims to independently investigate the gene expression data 
published by Cabrales-Orona et al. (2026) using differential expression 
analysis and original visualizations. My analysis focuses on comparing gene 
expression between wild-type plants and a transgenic line (Line 12) at matched 
time points to identify genes associated with heat tolerance. Below I outline 
the two main goals driving this exploration.

## Row Project Goals {height=40%}

### Column {width=50%}

#### Differential Expression Analysis
The primary goal of this dashboard is to identify genes that are 
differentially expressed between wild-type *Arabidopsis* and the transgenic 
Line 12 under heat shock conditions. Using DESeq2, I compare gene expression 
at the 1-hour heat shock time point to identify which genes are significantly 
up or downregulated in the transgenic line, and what biological processes 
those genes are involved in.

### Column {width=50%}

#### Heat Stress Response Across Time
The secondary goal is to explore how gene expression changes across multiple 
time points within each line. By visualizing expression patterns at 1 hour, 
5.5 hours, 11 hours, and after recovery, I aim to better understand how 
*Arabidopsis* adapts its transcriptional response to sustained heat stress, 
and whether the transgenic line responds differently over time compared to 
wild-type plants.

# Page 1 Data Set Up

## Column {width=30%}

```{r}
#| title: Data Visualization

data_sizes <- colSums(counts)
data_df <- data.frame(
  sample = names(data_sizes),
  total_counts = data_sizes
)

ggplot(data_df, aes(x = reorder(sample, total_counts), y = total_counts / 1000000)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Data per Sample",
    x = "Sample",
    y = "Total Counts (millions)"
  )

```

```{r}
#| title: Investigating Data Across Samples

log_counts <- log2(counts+1)
log_counts_long <- data.frame(
  expression = unlist(log_counts),
  sample = rep(colnames(log_counts), each = nrow(log_counts))
)

ggplot(log_counts_long, aes(x = expression, color = sample)) +
  geom_density() +
  labs(
    title = "Log2 Counts Across Samples",
    x = "Log2(Count + 1)",
    y = "Density"
  )

```


```{r}
#| title: Filtering and DESeq2 Object Set Up

#Creating a metadata table for use with DESeq2.
sample_info <- data.frame(
  row.names = colnames(counts),
  line = case_when(
    grepl("^WT", colnames(counts))  ~ "WT",
    grepl("^L12", colnames(counts)) ~ "Line12",
    grepl("^L13", colnames(counts)) ~ "Line13"
  ),
  condition = case_when(
    grepl("H", colnames(counts)) ~ "HeatShock",
    grepl("D", colnames(counts)) ~ "Recovery"
  ),
  timepoint = case_when(
    grepl("01", colnames(counts)) ~ "1hr",
    grepl("05", colnames(counts)) ~ "5.5hr",
    grepl("11", colnames(counts)) ~ "11hr",
    grepl("22", colnames(counts)) ~ "22hr"
  )
)
# printing table out, was feeling table might delete later.
sample_info

#Using the metadata table to create the DESeq object
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = sample_info,
  design    = ~ line
)
dds

#For use to better visualize the data
nrow(dds)

# filtering the data, drops low sample data from the data set for more robust results.
# This is important to remove genes we dont have enough data on to draw robust conclusions
# from, in other words it limits the sample bias, is I beleive the proper term.
dds <- dds[rowSums(counts(dds)) >= 10, ]

# genes after filtering
nrow(dds)
```


## Column {width=70%}

```{r}
#| title: PCA of the DESeq2 Object

# This normalizes the data so high-count genes don't dominate the PCA
vsd <- vst(dds, blind = TRUE)

# plot PCA
plotPCA(vsd, intgroup = c("line", "condition")) +
  labs(
    title = "PCA of All Samples",
    caption = "Shape indicates heat shock vs. recovery condition"
  )
```

# Page 2 Data Exploration and Visualization

## Column {width=30%}

```{r}
# Subset to just 1hr heatshock samples for clean comparison
samples_subset <- sample_info[
  sample_info$timepoint == "1hr" &
  sample_info$condition == "HeatShock" &
  sample_info$line %in% c("WT", "Line12"), ]

counts_subset <- counts[, rownames(samples_subset)]

dds_subset <- DESeqDataSetFromMatrix(
  countData = counts_subset,
  colData   = samples_subset,
  design    = ~ line
)

dds_subset$line <- relevel(dds_subset$line, ref = "WT")

# Filter low count genes
dds_subset <- dds_subset[rowSums(counts(dds_subset)) >= 10, ]

dds_subset <- DESeq(dds_subset)
res <- results(dds_subset, contrast = c("line", "Line12", "WT"))

summary(res)
```
## Column {width=70%}

```{r}
# Convert results to data frame
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

# Classify each gene as up, down, or not significant
res_df$significance <- case_when(
  res_df$padj < 0.05 & res_df$log2FoldChange >  1 ~ "Upregulated",
  res_df$padj < 0.05 & res_df$log2FoldChange < -1 ~ "Downregulated",
  TRUE ~ "Not Significant"
)

# Remove rows with NA adjusted p-values (DESeq2 filters some genes automatically)
res_df <- res_df[!is.na(res_df$padj), ]

# Plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c(
    "Upregulated"     = "firebrick",
    "Downregulated"   = "steelblue",
    "Not Significant" = "grey60"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(
    title    = "Volcano Plot: Line 12 vs Wild-Type (1hr Heat Shock)",
    x        = "Log2 Fold Change",
    y        = "-Log10 Adjusted P-value",
    color    = "Expression"
  )
```

# Page 3 Results and Conclusions

# Resources & Inspirations 

This project utilized resources primarily from two academic papers. If you are interested in anything you see on this dashboard look into the two papers cited below.

1. Cabrales-Orona G, et al. (2026). Grain amaranth genes coding for an RNA-binding and a small, unknown function protein, respectively, enhance thermotolerance when overexpressed in Arabidopsis thaliana. Physiol Mol Biol Plants, 32(1):59–80. PMID: 41675614
  1.2 The dataset used in the paper:
  NCBI Gene Expression Omnibus (GEO), accession GSE277320. https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE277320

2. Chen Y, Lun AT, Smyth GK. From reads to genes to pathways: differential expression analysis of RNA-Seq experiments using Rsubread and the edgeR quasi-likelihood pipeline. F1000Res. 2016 Jun 20;5:1438. doi: 10.12688/f1000research.8987.2. PMID: 27508061; PMCID: PMC4934518. 