---
title: "Exploration of wild-type and transgenic variants of Arabidopsis thaliana"
author: "Yves Vernooy"
format: 
  dashboard:
    theme: slate
    orientation: columns
embed-resources: true
---

```{r imports}
#| message: false
#| output: false

# Note: Some of these libraries are really annoying to install and
# need different commands beyond the typical install.packages()
library(knitr)
library(dplyr)
library(tidyverse)
library(pheatmap)
library(GEOquery)
library(ggplot2)
library(DESeq2)

# read in data
getGEOSuppFiles("GSE277320")
counts <- read.delim("GSE277320/GSE277320_CountsMatrix.txt.gz", row.names = 1)
```

# Introduction

# Page 1 Data Set Up

## Column {width=30%}

```{r}
#| title: Data Visualization

data_sizes <- colSums(counts)
data_df <- data.frame(
  sample = names(data_sizes),
  total_counts = data_sizes
)

ggplot(data_df, aes(x = reorder(sample, total_counts), y = total_counts / 1000000)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Data per Sample",
    x = "Sample",
    y = "Total Counts (millions)"
  )

```

```{r}
#| title: Investigating Data Across Samples

log_counts <- log2(counts+1)
log_counts_long <- data.frame(
  expression = unlist(log_counts),
  sample = rep(colnames(log_counts), each = nrow(log_counts))
)

ggplot(log_counts_long, aes(x = expression, color = sample)) +
  geom_density() +
  labs(
    title = "Log2 Counts Across Samples",
    x = "Log2(Count + 1)",
    y = "Density"
  )

```


```{r}
#| title: Filtering and DESeq2 Object Set Up

#Creating a metadata table for use with DESeq2.
sample_info <- data.frame(
  row.names = colnames(counts),
  line = case_when(
    grepl("^WT", colnames(counts))  ~ "WT",
    grepl("^L12", colnames(counts)) ~ "Line12",
    grepl("^L13", colnames(counts)) ~ "Line13"
  ),
  condition = case_when(
    grepl("H", colnames(counts)) ~ "HeatShock",
    grepl("D", colnames(counts)) ~ "Recovery"
  ),
  timepoint = case_when(
    grepl("01", colnames(counts)) ~ "1hr",
    grepl("05", colnames(counts)) ~ "5.5hr",
    grepl("11", colnames(counts)) ~ "11hr",
    grepl("22", colnames(counts)) ~ "22hr"
  )
)
# printing table out, was feeling table might delete later.
sample_info

#Using the metadata table to create the DESeq object
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = sample_info,
  design    = ~ line
)
dds

#For use to better visualize the data
nrow(dds)

# filtering the data, drops low sample data from the data set for more robust results.
# This is important to remove genes we dont have enough data on to draw robust conclusions
# from, in other words it limits the sample bias, is I beleive the proper term.
dds <- dds[rowSums(counts(dds)) >= 10, ]

# genes after filtering
nrow(dds)
```


## Column {width=70%}

```{r}
#| title: PCA of the DESeq2 Object

# This normalizes the data so high-count genes don't dominate the PCA
vsd <- vst(dds, blind = TRUE)

# plot PCA
plotPCA(vsd, intgroup = c("line", "condition")) +
  labs(
    title = "PCA of All Samples",
    caption = "Shape indicates heat shock vs. recovery condition"
  )
```

# Page 2 Data Exploration and Visualization

## Column {width=30%}

```{r}
# Subset to just 1hr heatshock samples for clean comparison
samples_subset <- sample_info[
  sample_info$timepoint == "1hr" &
  sample_info$condition == "HeatShock" &
  sample_info$line %in% c("WT", "Line12"), ]

counts_subset <- counts[, rownames(samples_subset)]

dds_subset <- DESeqDataSetFromMatrix(
  countData = counts_subset,
  colData   = samples_subset,
  design    = ~ line
)

dds_subset$line <- relevel(dds_subset$line, ref = "WT")

# Filter low count genes
dds_subset <- dds_subset[rowSums(counts(dds_subset)) >= 10, ]

dds_subset <- DESeq(dds_subset)
res <- results(dds_subset, contrast = c("line", "Line12", "WT"))

summary(res)
```
## Column {width=70%}

```{r}
# Convert results to data frame
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

# Classify each gene as up, down, or not significant
res_df$significance <- case_when(
  res_df$padj < 0.05 & res_df$log2FoldChange >  1 ~ "Upregulated",
  res_df$padj < 0.05 & res_df$log2FoldChange < -1 ~ "Downregulated",
  TRUE ~ "Not Significant"
)

# Remove rows with NA adjusted p-values (DESeq2 filters some genes automatically)
res_df <- res_df[!is.na(res_df$padj), ]

# Plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c(
    "Upregulated"     = "firebrick",
    "Downregulated"   = "steelblue",
    "Not Significant" = "grey60"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(
    title    = "Volcano Plot: Line 12 vs Wild-Type (1hr Heat Shock)",
    x        = "Log2 Fold Change",
    y        = "-Log10 Adjusted P-value",
    color    = "Expression"
  )
```

# Page 3 Results and Conclusions

# Resources & Inspirations 

This project utilized resources primarily from two academic papers. If you are interested in anything you see on this dashboard look into the two papers cited below.

1. Cabrales-Orona G, et al. (2026). Grain amaranth genes coding for an RNA-binding and a small, unknown function protein, respectively, enhance thermotolerance when overexpressed in Arabidopsis thaliana. Physiol Mol Biol Plants, 32(1):59–80. PMID: 41675614
  1.2 The dataset used in the paper:
  NCBI Gene Expression Omnibus (GEO), accession GSE277320. https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE277320

2. Chen Y, Lun AT, Smyth GK. From reads to genes to pathways: differential expression analysis of RNA-Seq experiments using Rsubread and the edgeR quasi-likelihood pipeline. F1000Res. 2016 Jun 20;5:1438. doi: 10.12688/f1000research.8987.2. PMID: 27508061; PMCID: PMC4934518. 