---
title: "Exploration of wild-type and transgenic variants of Arabidopsis Thaliana"
author: "Yves Vernooy"
format: 
  dashboard:
    theme: slate
    orientation: columns
embed-resources: true
---

```{r imports}
#| message: false
#| output: false

# Note: Some of these libraries are really annoying to install and
# need different commands beyond the typical install.packages()
library(knitr)
library(dplyr)
library(tidyverse)
library(pheatmap)
library(GEOquery)
library(ggplot2)
library(DESeq2)

# read in data
getGEOSuppFiles("GSE277320")
counts <- read.delim("GSE277320/GSE277320_CountsMatrix.txt.gz", row.names = 1)
```

# Introduction

## Row {height=60%}

::: {.card title="Project Aims and Data Set Introduction"}
This project is a Quarto dashboard exploring differential gene expression 
in *Arabidopsis thaliana* under heat stress conditions, using a publicly 
available RNA-seq dataset from NCBI's Gene Expression Omnibus (GEO accession: 
GSE277320). This dataset was generated by Cabrales-Orona et al. (2026) and 
contains gene expression profiles from wild-type *Arabidopsis* plants and 
transgenic lines overexpressing RNA-binding proteins originally identified in 
grain amaranth (*Amaranthus hypochondriacus*), a heat-tolerant crop plant. 
All 30 samples were collected under heat shock conditions across five time 
points, with two biological replicates per group.

This project aims to independently investigate the gene expression data 
published by Cabrales-Orona et al. (2026) using differential expression 
analysis and original visualizations. My analysis focuses on comparing gene 
expression between wild-type plants and a transgenic line (Line 12) at matched 
time points to identify genes associated with heat tolerance. Below I outline 
the two main goals driving this exploration.
:::

## Row {height=40%}

::: {.card title="Differential Expression Analysis"}
The primary goal of this dashboard is to identify genes that are 
differentially expressed between wild-type *Arabidopsis* and the transgenic 
Line 12 under heat shock conditions. Using DESeq2, I compare gene expression 
at the 1-hour heat shock time point to identify which genes are significantly 
up or downregulated in the transgenic line, and what biological processes 
those genes are involved in.
:::

::: {.card title="Heat Stress Response Across Time"}
The secondary goal is to explore how gene expression changes across multiple 
time points within each line. By visualizing expression patterns at 1 hour, 
5.5 hours, 11 hours, and after recovery, I aim to better understand how 
*Arabidopsis* adapts its transcriptional response to sustained heat stress, 
and whether the transgenic line responds differently over time compared to 
wild-type plants.
:::

# Data Set Up

## Column {width=40%}
::: {.card title="Data Visualization"}
```{r}
data_sizes <- colSums(counts)
data_df <- data.frame(
  sample = names(data_sizes),
  total_counts = data_sizes
)

ggplot(data_df, aes(x = reorder(sample, total_counts), y = total_counts / 1000000)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Data per Sample",
    x = "Sample",
    y = "Total Counts (millions)"
  )
```

Here we start our data exploration with visualizing the total sequencing depth across all 30 samples of *Arabidopsis thaliana*. This tells us something important there is a drop off of data points for some lines and most notably a harsh drop off for two lines specifically WT03D2 and WT03D1.
:::

::: {.card title="Investigating Data Across Samples"}
```{r}
log_counts <- log2(counts+1)
log_counts_long <- data.frame(
  expression = unlist(log_counts),
  sample = rep(colnames(log_counts), each = nrow(log_counts))
)

ggplot(log_counts_long, aes(x = expression, color = sample)) +
  geom_density() +
  labs(
    title = "Log2 Counts Across Samples",
    x = "Log2(Count + 1)",
    y = "Density"
  )

```
The density plot above shows the distribution of log2-transformed counts across all 30 samples. The large spike near zero reflects the high proportion of genes with zero or close to zero expression which is a expected as most genes are not actively expressed in any given tissue or condition. The close overlap between all 30 sample curves suggests quality data.
:::
```{r}

#Creating a metadata table for use with DESeq2.
sample_info <- data.frame(
  row.names = colnames(counts),
  line = case_when(
    grepl("^WT", colnames(counts))  ~ "WT",
    grepl("^L12", colnames(counts)) ~ "Line12",
    grepl("^L13", colnames(counts)) ~ "Line13"
  ),
  condition = case_when(
    grepl("H", colnames(counts)) ~ "HeatShock",
    grepl("D", colnames(counts)) ~ "Recovery"
  ),
  timepoint = case_when(
    grepl("01", colnames(counts)) ~ "1hr",
    grepl("05", colnames(counts)) ~ "5.5hr",
    grepl("11", colnames(counts)) ~ "11hr",
    grepl("22", colnames(counts)) ~ "22hr"
  )
)

#Using the metadata table to create the DESeq object
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = sample_info,
  design    = ~ line
)

# filtering the data, drops low sample data from the data set for more robust results.
# This is important to remove genes we don't have enough data on to draw robust conclusions
# from, in other words it limits the sample bias, is I believe the proper term.
dds <- dds[rowSums(counts(dds)) >= 10, ]
```

## Column {width=60%}
::: {.card title="PCA of the DESeq2 Object"}
```{r}
# This normalizes the data so high-count genes don't dominate the PCA
vsd <- vst(dds, blind = TRUE)

# plot PCA
plotPCA(vsd, intgroup = c("line", "condition")) +
  labs(
    title = "PCA of All Samples",
    caption = "Shape indicates heat shock vs. recovery condition"
  )
```
Here we are plotting the PCA of all the samples in the dataset after we filtered the data. This shows us the amount of data that each line in heat shock or recovery contributes to the variance in the data.
:::

# Data Analysis and Vizualization

## Column {width=30%}
::: {.card title="Differential Expression Analysis"}
```{r}
# Subset to just 1hr heatshock samples for clean comparison
samples_subset <- sample_info[
  sample_info$timepoint == "1hr" &
  sample_info$condition == "HeatShock" &
  sample_info$line %in% c("WT", "Line12"), ]

counts_subset <- counts[, rownames(samples_subset)]

dds_subset <- DESeqDataSetFromMatrix(
  countData = counts_subset,
  colData   = samples_subset,
  design    = ~ line
)

dds_subset$line <- relevel(dds_subset$line, ref = "WT")

# Filter low count genes
dds_subset <- dds_subset[rowSums(counts(dds_subset)) >= 10, ]

dds_subset <- DESeq(dds_subset)
res <- results(dds_subset, contrast = c("line", "Line12", "WT"))

summary(res)
```

To identify genes that respond differently to heat stress between wild-type 
*Arabidopsis* and the transgenic Line 12, we use DESeq2, a statistical 
framework designed specifically for RNA sequence count data. Our analysis subsets 
the data to only the 1 hour heat shock time point which gives us the cleanest 
treated vs. control comparison. Wild-type is set as the reference group, so 
fold changes represent expression in Line 12 relative to WT. Genes with an 
adjusted p-value below 0.05 and a log2 fold change greater than 1 in either 
direction are considered statistically significant.
:::
## Column {width=70%}

::: {.card title="Volcano Plot: Line 12 vs Wild-Type"}
```{r}
# Convert results to data frame
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

# Classify each gene as up, down, or not significant
res_df$significance <- case_when(
  res_df$padj < 0.05 & res_df$log2FoldChange >  1 ~ "Upregulated",
  res_df$padj < 0.05 & res_df$log2FoldChange < -1 ~ "Downregulated",
  TRUE ~ "Not Significant"
)

# Remove rows with NA adjusted p-values (DESeq2 filters some genes automatically)
res_df <- res_df[!is.na(res_df$padj), ]

# Plot
ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c(
    "Upregulated"     = "firebrick",
    "Downregulated"   = "steelblue",
    "Not Significant" = "grey60"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(
    title    = "Volcano Plot: Line 12 vs Wild-Type (1hr Heat Shock)",
    x        = "Log2 Fold Change",
    y        = "-Log10 Adjusted P-value",
    color    = "Expression"
  )
```
The volcano plot visualizes the results of the differential expression 
analysis. Each point represents a single gene. The x-axis shows the log2 
fold change, how much higher or lower a gene's expression is in Line 12 
compared to wild-type. The y-axis shows statistical significance as 
-log10 of the adjusted p-value, meaning points higher up are more 
statistically significant. Red points in the upper right are significantly 
upregulated in Line 12, blue points in the upper left are significantly 
downregulated, and grey points did not meet our significance thresholds. 
The dashed lines mark the fold change (±1) and significance (padj = 0.05) 
thresholds.
:::

# Time Course Analysis

## Column {width=30%}
::: {.card title="Top Genes"}
```{r}
# Subset to WT and Line12, heat shock only (exclude recovery samples)
samples_timecourse <- sample_info[
  sample_info$condition == "HeatShock" &
  sample_info$line %in% c("WT", "Line12"), ]

counts_timecourse <- counts[, rownames(samples_timecourse)]

dds_time <- DESeqDataSetFromMatrix(
  countData = counts_timecourse,
  colData   = samples_timecourse,
  design    = ~ line + timepoint
)

dds_time$line <- relevel(dds_time$line, ref = "WT")
dds_time <- dds_time[rowSums(counts(dds_time)) >= 10, ]
dds_time <- DESeq(dds_time)

# Variance stabilizing transformation for visualization
vsd_time <- vst(dds_time, blind = FALSE)

# Take top 40 most significant DE genes from your earlier analysis
top_genes <- res_df |>
  filter(significance != "Not Significant") |>
  arrange(padj) |>
  head(40) |>
  pull(gene)

print(top_genes)
```
Here we construct a new DESeq2 object using only the WT and Line 12 samples collected during active heat shock, with recovery samples excluded. differing from the volcano plot analysis we incorporate both line and time point as variables, allowing us to capture expression dynamics across 1, 5.5, and 11 hours of heat stress. We then extract the top 40 most statistically significant genes identified in the previous analysis as the basis for our time course visualization.
:::
## Column {width=70%}
::: {.card title="temp"}
```{r}
# Extract normalized expression for top genes
mat <- assay(vsd_time)[top_genes, ]

# Scale each gene (row) so colors show relative expression
mat_scaled <- t(scale(t(mat)))

# Build annotation for columns showing line and timepoint
annotation_col <- data.frame(
  row.names = colnames(mat),
  Line      = samples_timecourse$line,
  Timepoint = samples_timecourse$timepoint
)

pheatmap(
  mat_scaled,
  annotation_col  = annotation_col,
  cluster_cols    = FALSE,   # keep time order meaningful
  cluster_rows    = TRUE,    # cluster genes by expression pattern
  show_rownames   = TRUE,
  show_colnames   = FALSE,
  color           = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
  main            = "Top DE Genes: WT vs Line 12 Across Heat Shock Time Points"
)
```
The heatmap above visualizes the expression patterns of the top 40 differentially expressed genes across all heat shock time points in WT and Line 12 plants. Each row represents a gene and each column represents a sample, with red indicating higher relative expression and blue indicating lower relative expression. Genes are clustered by similarity in expression pattern, revealing groups of genes that respond to heat stress in a coordinated manner. By comparing columns across time points and between lines, we can see whether the transcriptional differences between wild-type and Line 12 are consistent throughout heat exposure or change as stress is sustained.
:::

# Summary & Resources 

## Row {height=70%}
::: {.card title="Summary and Findings"}

:::
## Row {height=30%}
::: {.card title="Resources and Citations"}
This project utilized resources primarily from two academic papers. If you are interested in anything you see on this dashboard look into the two papers cited below.

1. Cabrales-Orona G, et al. (2026). Grain amaranth genes coding for an RNA-binding and a small, unknown function protein, respectively, enhance thermotolerance when overexpressed in Arabidopsis thaliana. Physiol Mol Biol Plants, 32(1):59–80. PMID: 41675614
  1.2 The dataset used in the paper:
  NCBI Gene Expression Omnibus (GEO), accession GSE277320. https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE277320

2. Chen Y, Lun AT, Smyth GK. From reads to genes to pathways: differential expression analysis of RNA-Seq experiments using Rsubread and the edgeR quasi-likelihood pipeline. F1000Res. 2016 Jun 20;5:1438. doi: 10.12688/f1000research.8987.2. PMID: 27508061; PMCID: PMC4934518. 
:::